Full Brain Blood-Oxygen-Level-Dependent Signal Parameter Estimation Using Particle Filters

### 2010-08-23 doc/paper/main.tex
added simulation stuff

\begin{abstract}
Traditional methods of analyzing FMRI images use a linear combination of
just a few static regressors. This works demonstrates an alternative
approach using a physiologically inspired nonlinear model. By using a 
particle filter to optimize the model parameters, the computation time
is kept below a minute per voxel without requiring a linearization 
of the noise in the state
variables. The activation results show regions similar those found in 
SPM; however, there are some notable regions not detected by 
SPM. Though the parameters selected by the particle filter based approach
are more than sufficient to predict the BOLD response,
more model constraints are needed to uniquely identify a single set
of parameters. This ill-posed nature explains the large discrepancies
found in other research that attempted to characterize the model parameters.
For this reason the final distribution of parameters is more medically relevant
than a single estimate. Because the output of the particle filter is 
a full posterior probability, the reliance on the mean to estimate 
parameters is unnecessary. This work presents
not just a viable alternative to the traditional method of detecting
activation, but an extensible technique of estimating the joint probability
distribution of the BOLD parameters.
\end{abstract}

@@

Plugging \autoref{eq:ImportanceDensity} and the result of \autoref{eq:UpdateBayes}
into \autoref{eq:weightfunc} leads to:
\begin{eqnarray}
w^i_k & \propto & \frac{P(y_k| x^i_k)  P(x^i_{0:k-1} | y_{0:k-1})}
                         {q(x^i_{0:k-1} | y_{0:k-1})} \nonumber \\
& \propto & w^i_{k-1}P(y_k| x_k) 
\label{eq:weightevolve}
\end{eqnarray}

@@

For infinite particles this new distribution will match the old.
Unfortunately, in practice since the support is
limited to the original particles, the number of \emph{unique} 
particles can only decrease.
This effect, called particle impoverishment, can result in 
excessive quantization errors in the final distribution. 
To prevent particle impoverishment, instead of sampling from the
discrete distribution, a smoothing kernel is applied.
Because the distribution is continuous, particle impoverishment
is prevented. This regularization process is defined as:
\begin{equation}
x_i = x_i + h\sigma \epsilon
\end{equation}
Where $h$ is an optional bandwidth, $\sigma$ is the standard 
deviation such that $\sigma \sigma^T = cov(x)$
and $\epsilon$ is drawn from the chosen kernel. In this case
a Gaussian kernel was used, and the bandwidth was set to 1
(see \autoref{alg:RegResampling}). All the 
pseudocode may be found in \autoref{sec:algorithms},
including the complete regularized particle filter used
(Algorithm \autoref{alg:RegularizedParticleFilter}).

\subsection{Simulation}
\label{sec:MethodsSim}
\subsection{Data Acquisition}
\label{sec:MethodsData}

\section{Results}
\label{sec:Results}
\begin{itemize}
\item Simulation Results - Possum Only
\begin{itemize}
    \item Histogram of parameters for each region
    \item Final histogram of one run?
    \item Comparison of MI with SNR
\end{itemize}
\item FMRI Results 
\begin{itemize}
    \item Activation maps comparing M.I. with SPM
\end{itemize}
\end{itemize}

>>>>

\begin{abstract}
Traditional methods of analyzing FMRI images use a linear combination of
just a few static regressors. This work demonstrates an alternative
approach using a physiologically inspired nonlinear model. By using a 
particle filter to optimize the model parameters, the computation time
is kept below a minute per voxel without requiring a linearization 
of the noise in the state
variables. The activation results show regions similar to those found in 
SPM; however, there are some notable regions not detected by 
SPM. Though the parameters selected by the particle filter based approach
are more than sufficient to predict the BOLD response,
more model constraints are needed to uniquely identify a single set
of parameters. This ill-posed nature explains the large discrepancies
found in other research that attempted to characterize the model parameters.
For this reason the final distribution of parameters is more medically relevant
than a single estimate. Because the output of the particle filter is 
a full posterior probability, the reliance on the mean to estimate 
parameters is unnecessary. This work presents
not just a viable alternative to the traditional method of detecting
activation, but an extensible technique of estimating the joint probability
distribution function of the BOLD parameters.
\end{abstract}

@@

Plugging \autoref{eq:ImportanceDensity} and the result of \autoref{eq:UpdateBayes}
into \autoref{eq:weightfunc} leads to:
\begin{eqnarray}
w^i_k & \propto & \frac{P(y_k| x^i_k)  P(x^i_{0:k-1} | y_{0:k-1})}
                         {q(x^i_{0:k-1} | y_{0:k-1})} \nonumber \\
& \propto & w^i_{k-1}P(y_k| x^i_k) 
\label{eq:weightevolve}
\end{eqnarray}

@@

For infinite particles this new distribution will match the old.
Unfortunately, in practice since the support is
limited to the original particles, the number of \emph{unique} 
particles can only decrease.
This effect, called particle impoverishment, can result in 
excessive quantization errors in the final distribution. 
To prevent particle impoverishment, instead of sampling from the
discrete distribution, a smoothing kernel is applied.
Because the distribution is continuous, particle impoverishment
is prevented. This regularization process is defined as:
\begin{equation}
x_i = x_i + h\sigma \epsilon
\end{equation}
Where $h$ is an optional bandwidth, $\sigma$ is the standard 
deviation such that $\sigma \sigma^T = cov(x)$
and $\epsilon$ is drawn from the chosen kernel. In this case
a Gaussian kernel was used, and the bandwidth was set to 1
(see \autoref{alg:RegResampling}). All the 
pseudocode may be found in \autoref{sec:algorithms},
including the complete regularized particle filter used
(Algorithm \autoref{alg:RegularizedParticleFilter}).
\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c |}
\hline
Parameter & Distribution & $\mu$ & $\sigma$ \\
\hline
$\tau_0$ & Gamma & .98 & .25 \\
$\alpha$ & Gamma & .33 & .045\\
$E_0$    & Gamma & .34 & .03  \\
$V_0$    & Gamma & .04 & .03 \\
$\tau_s$ & Gamma & 1.54  & .25\\
$\tau_f$ & Gamma & 2.46  & .25\\
$\epsilon$ & Gamma & .7  & .6 \\
\hline
\end{tabular}
\caption{Prior distributions used in the particle filter.}
\label{tab:Prior}
\end{table}
\autoref{eq:weightevolve} still depends on knowing
$P(y_k| x^i_k)$. Ideally this would exactly match the
measurement error distribution of the system; however,
even if this distribution is incorrect, the solution will
still converge. For this work it was assumed that
\begin{equation}
P(y_k | x^i_k) = g(X_k, \nu_y) = g(X_k) + \nu_y\text{, }\nu_y \sim N(0, 0.005^2)
\end{equation}
A standard deviation of $0.005$ performed well in the general case, primarily
because the BOLD signal tends to be in the range of $0.01-0.05$, (1\%-5\%).
Additionally, 1400 Euler integration points were used per second, 
the system was assumed to start at rest. The system was initialized
with 16,000 particles, but during the first resampling, the number
of particles was dropped to 1,000. To prevent over-smoothing, resampling
was only performed when $\N_{eff}$ (\autoref{eq:neff}) dropped below
$50$ for two consecutive measurements.

\subsection{Preprocessing}
The BOLD response has been
extensively studied and despite minor discrepancies, the cause of the BOLD
signal is well known. However, FMRI has an extremely low SNR. 
A particularly difficult form of noise present 
in FMRI is a low frequency
drift, often characterized as a Wiener process \cite{Riera2004}.
Though not present in all regions, as many as ten to fifteen percent
of voxels can be affected, thus it is prevalent 
enough to cause significant inference problems \cite{Tanabe2002, Smith2007}. 
Though the cause is still not
clear it is clear that this drift signal is not solely
due to a physiological effects, given its presence in cadavers and phantoms
\cite{Smith1999}. The problem mandates the use of a high pass 
filter \cite{Smith2007}.  In a comparison of high pass filters,
Tanabe et al.  showed that generally subtracting off
a spline improved the SNR the most \cite{Tanabe2002}.

To prevent transients at the beginning of the FMRI run, the first 9
measurements (18.9 seconds) were dropped.
One knot was used for every 20 time points (at a TR of 2.1), with the first
knot and last knot taking half as many points as the rest of the knots.
The median level of each group of measurements was calculated, and the
knot placed at the average time of the group. After subtracting this spline
from the timeseries, the resulting timeseries was divided by the original
sample mean. To compensate for the BOLD signal not generally being zero-mean,
the Median-Absolute-Deviation was added to the final signal (\autoref{eq:mad}). 
\begin{equation}
y_{\text{gain}, 0:K} = 1.4826\underset{i=0:K}{\text{median}}(y_i - \text{median}(y_{0:K}))
\label{eq:mad}
\end{equation}

\subsection{Simulation}
\label{sec:MethodsSim}
Single voxel simulations were performed by propagating the BOLD
model forward through time with 2000 integration points per
second, then adding I.I.D. Gaussian noise and a Wiener process
with I.I.D. steps. 
Full volume simulation were performed using 
a modified version of the FSL tool POSSUM. The modifications allowed
activation levels to be driven by spatially dependent BOLD parameter
sets. Due to the modifications, the SNR configuration was set to 20,
although actual SNR's ranged from .5 to 1.5 in active areas.

\subsection{Data Acquisition}
\label{sec:MethodsData}

\section{Results}
\label{sec:Results}
\subsection{Simulation}
The final
histogram for a single active region, with a single set of true 
parameters is shown in \autoref{fig:slicesim_hist2}.

%parameters:
%&1.151&  0.353& 0.380& 0.026& 1.98&  2.333& 1.645 \\
\begin{figure*}[!t] %top left
\centering
\includegraphics[clip=true,trim=2.5cm 2cm 2cm 1cm,width=15cm]{images/slicesim_hist2}
\caption{Histogram of estimated parameters. Average SNR in the region was 
$0.97$ and Mutual Information between simulation and BOLD estimate was greater 
than $0.15$. TODO: add actual error.}
\label{fig:slicesim_hist2}
\end{figure*}

\begin{itemize}
\item Long Single Result: Correlation Matrix, error in estimate
\item Possum Results: Histogram of parameters for a region
\item FMRI Results: Activation maps comparing M.I. with SPM
